---
title: "Gov 50 Final Project"
author: Hannah Ahn
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Milestone 4


```{r, echo= FALSE}
#Setup!
library(tidyverse)
library(ggplot2)
```

```{r, echo= FALSE}
master_data <- read_csv("data/gov50finaldata.csv")

master_data_plot <- ggplot(data = master_data, mapping = aes(x = `Gun Law Strength 2020`, y = `Suicide Rate 2020`, label = `State`)) +
  geom_point() +
  geom_text() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Gun Law Strength vs Suicide Rate by Year",
       x = "Gun Law Weakness",
       y = "Suicide Rate per 100k")

regression <- lm(`Suicide Rate 2020` ~ `Gun Law Strength 2020`, data = master_data)

master_data_plot

summary(regression)

```

# Milestone 3
```{r, echo=FALSE}
library(tidyverse)
library(ggplot2)
sui_data <- read_csv("data/suicide_death_rates.csv")
overdose_data <- read_csv("data/overdose_death_rate.csv")

#Formatting data!

sui_data1 <- subset(sui_data, select = -c(INDICATOR, UNIT_NUM, STUB_LABEL_NUM, YEAR_NUM, STUB_NAME_NUM, AGE_NUM, FLAG)) |> 
  filter(`STUB_NAME` == "Sex and race", YEAR >= 1999, UNIT == "Deaths per 100,000 resident population, age-adjusted") |> 
  mutate(gender = if_else(STUB_LABEL == "Female: American Indian or Alaska Native" |STUB_LABEL == "Female: Asian or Pacific Islander" | STUB_LABEL == "Female: Black or African American" | STUB_LABEL ==  "Female: White", "Female", "Male"))

sui_plot <- ggplot(data = sui_data1, mapping = aes(x = YEAR, y = ESTIMATE)) + geom_line(mapping = aes(color = STUB_LABEL)) + facet_wrap(~ gender, ncol = 4) + xlab("Year") + ylab("Estimated Number of Deaths per 100,000") + labs(title = "Suicides by Race and Gender", color = "Gender and Race")



#Doing the same to overdose data! 

overdose_data1 <- subset(overdose_data, select = -c(INDICATOR, PANEL_NUM, UNIT, UNIT_NUM, STUB_LABEL_NUM, YEAR_NUM, STUB_NAME_NUM, AGE_NUM, FLAG)) |> 
  filter(`STUB_NAME` == "Sex and race", `PANEL` == "All drug overdose deaths")|>
  mutate(gender = if_else(STUB_LABEL == "Female: American Indian or Alaska Native" |STUB_LABEL == "Female: Asian or Pacific Islander" | STUB_LABEL == "Female: Black or African American" | STUB_LABEL ==  "Female: White", "Female", "Male"))

overdose_plot <- ggplot(data = overdose_data1, mapping = aes(x = YEAR, y = ESTIMATE)) + geom_line(mapping = aes(color = STUB_LABEL)) + facet_wrap(~ gender, ncol = 4) + xlab("Year") + ylab("Estimated Number of Deaths per 100,000") + labs(title = "Overdoses by Race and Gender", color = "Gender and Race")

sui_plot
overdose_plot
```

## Milestone 2 

Does firearm legislation affect the incidence of successful suicide? In this study, I plan to examine the extent to which firearm legislation lowers (or raises) the incidence of successful suicide. For the purposes of this study, I will use data from the Giffords Law Center. My explanatory variable of interest is the change in Gun Law Grade as determined by the GIfford Law institute. My explanatotry variable is masured from the Gifford Law Institute and depends on its ability to correctly classify and rank different states' gun law strengths. If I observe that suicide rates increase after states' gun law strength increases, this would provide support for my hypothesis. If I observe that suicide rates stay the same or decrease after states' gun law strength increases, this would provide evidence against my hypothesis. 

## Milestone 1

I am interested in examining whether firearm legislation affects incidence of suicide in the United states. 
